Settler colonial projects explain variation in US state formation in the early 20th c.

### 2024-01-04 ms.Rmd
ff

---
title: "Settler colonial projects and variation in paths to building the welfare and carceral state"
author: "Frank Edwards"
date: "last updated `r Sys.Date()`"
output: bookdown::html_document2
abstract: "Place and time variation in settler colonial projects are related to the kinds of states that settler elites build."
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)

options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
```

```{r read, include = F, message = F, cache = T}

### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")

## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct())

dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct())

source("models.R")
```

@@

This paper proposes a comparative quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects). We show that spatial and temporal variation in modes of settler colonial extraction and exploitation are closely related to trajectories of state formation across places.

At a sub-national level, settler colonial regimes seek to address a range of problems inherent to their nation-building project: the dispossession of indigenous lands, the erasure of indigenous peoples, the exploitation of racially marginalized laborers, and the reproduction and expansion of the settler polity. We show that the historical geography of these problems of settler colonial governance were closely related to the policy infrastructure to manage social life that governments develop.

<!-- Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it? -->

Following Wolfe, DuBois, etc, variation in the interests of elites across places (material + cultural) drives variation in processes of racialization across places. The state is a key vector through which racialization is implemented. Early 20th c. saw rapid development and expansion of social state. Can we observe differences in social state developmental trajectories as a function of variation in elite interests in particular racial projects? Are regional differences in social policy regimes a function of the social, demographic, and economic interests of regional elites?

<!-- Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native. Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion. Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines. -->

<!-- Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development. This can lock in particular racial regimes over time as social policy does work on social groups. -->

<!-- 1. Frontier settler polity -> elite interest in disposession / assimiatilion -> aggressive paternalistic welfare state for AIAN, social democracy for settlers -->
<!-- 2. Jim Crow -> elite interest in exploitation and exclusion -> elite paternalism, malign neglect, hard coercion -->
<!-- 3. Metropole settler polity -> elite interests in labor discipline / immig assimilation -> progressive paternalism, strong coercive welfare state for W, hard coercion and contested welfare for Blk -->

<!-- IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.  -->

<!-- # What is already known about settler colonization and state formation in late 19th and Early 20th c.  -->

<!-- ## Frontier -->

<!-- review here -->

<!-- ## Jim Crow -->

<!-- review here -->

<!-- ## Metropole  -->

<!-- review here -->

# Methods

*We evaluate whether a set of policy outcomes {employment, infrastructure} are related to a set of demographic and economic predictors {population composition, economic activity, ownership regime} at the state and year-levels.*

To do so, we rely on historical data from the US Census accessed via IPUMS (Ruggles et al), from the Bureau of Land Management (Archived by the University of Nebraska), from the US Department of the Interior (CITE BS REPORT), and from records our team digitized from the Commissioner on Indian Affairs (CITE).

<!-- Less interested in causal identification, more interested in understanding spatial patterns and trajectories. We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data. Move from specific to general.  -->

Generally our approach demonstrates suggests that state formation in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.

<!-- negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1 -->

<!-- positive case: white immigrants -> more social workers in t1 -->

<!-- here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.  -->

<!-- ## NOTES ON GEOGRAPHIES and MEASUREMENT -->

<!-- TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION. CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION. SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.  -->

<!-- STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).  -->

<!-- geography itself endogenous to settler colonization, this is tricky! -->

## Sources of data

Census (pop, employment, group quarters); ARCIA; Interior; BLM; US Forest Service.

<!-- CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho... -->

<!-- we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.  -->

<!-- let's look at how it played out in the progressive era across states and regions. we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression. -->

<!-- First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita? -->

We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states. A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.). We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.

### measures

All measures are computed at the state-year level.

EMPLOYMENT:

- teachers (1900 - 1930 100%)
- cops (1900 - 1930 100%)
- social workers (1900 - 1930 100%)

POPULATION:

- total (1900 - 1930 1%)
- children (1900 - 1930 1%)
- white immigrants in *urban households* (1900 - 1930 1%)
- Black population living in tenant farm households (1900 - 1930 1%)
- Population in group quarters (correctional institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
- Population in group quarters (mental institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
- Population in group quarters (institutions for the elderly, handicapped, and poor) (1900 100%, 1910 1.4%, 1920 - 1930 100%)

LAND:

- cumulative acres parceled through the Morill Act (Robert Lee, ``Morrill Act of 1862 Indigenous Land Parcels Database,'' High Country News, 2020)
- cumulative acres parceled through the Homestead Act (``Homesteads,'' Bureau of Land Management of the U.S. Department of the Interior, 1962. digitized by Edwards, Friefeld and and Wingo 2017)

Model is

\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \delta_t + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} \]

```{r fig.cap = "demographic map of US in 1900 - Black, White immig, white non-immig"}
plot_dat<-pop %>% 
  filter(year == 1900) %>% 
  mutate(white_non_immig = white_non_immig / (white_non_immig + white_immig),
         white_immig = 1 - white_non_immig,
         black = blk / tot_pop) %>% 
  select(STUSPS, white_immig, white_non_immig, black) %>% 
  pivot_longer(cols = white_immig:black) %>% 
  mutate(name = case_when(
    name == "black" ~ "Black / \ntotal population",
    name == "white_immig" ~ "White immigrant/\ntotal white population",
    name == "white_non_immig" ~ "White non-immigrant/\ntotal white population"
  ))

st %>% 
  left_join(plot_dat) %>% 
  filter(!is.na(value)) %>% 
  ggplot(aes(fill = value)) + 
  geom_sf() + 
  facet_wrap(~name, ncol = 3) + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "")
```

This map shows key features of the demography of the US in 1900 that are closely related to the geography of settler colonialism in the early 20th century US.

BLACK POPS IN JIM CROW SOUTH

US BORN WHITE POPS IN SOUTH, NEW ENGLAND, RURAL PARTS OF INNER MIDWEST

FOREIGN BORN WHITE POPS THROUGHOUT CITIES, WESTERN FRONTIER

FIG 2, 3 - native disposession

```{r}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

@@

ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year",
       title = "Post-1784 Indigenous disposession",
       subtitle = "Cessions to US federal government through 1894") 
```

@@

```{r fig.cap = "Disposession of Native lands in the 19th c."}
p1<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = cumulative_ceded / 1e6)) + 
  labs(x = "Year",
       y = "Acres (millions)") + 
  lims(y = c(0, 150)) + 
  facet_wrap(region~division) + 
  labs(subtitle = "1. Total land cessions")

@@

```{r fig.cap = "Two mechanisms for the distribution of disposessed Native Lands to settlers"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1900) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Lands patented under Morill Act (1862) by 1900") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Lands patented under Homestead Act (1862) by 1900") + 
  theme(legend.position = "bottom")

@@

FIGURE - by employment type, facet ~ year

```{r}
pdat<-pop %>% 
  left_join(st)
# could use children in denominator here
# keeping as total population for comparison 
ggplot(pdat,
       aes(fill = teachers / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Teachers per\n1,000 population")

ggplot(pdat,
       aes(fill = police / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Police per\n1,000 population")

ggplot(pdat,
       aes(fill = socwork / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Social Workers per\n1,000 population")

@@

FIG 2 - employment maps.

INST MAPS

```{r}
ggplot(pdat, aes(fill = inst_correctional / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  theme_void() + 
  scale_fill_viridis_c()

ggplot(pdat, aes(fill = inst_mental / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  theme_void() + 
  scale_fill_viridis_c()

ggplot(pdat, aes(fill = inst_poor_disabled / tot_pop * 1e3)) + 
  geom_sf(aes(geometry = geometry)) + 
  facet_wrap(~year) + 
  theme_void() + 
  scale_fill_viridis_c()
```

THEN MODELS

DISPOSESSION MEASURES - YEAR OF LAST TREATY SIGNED, ACRES PATENTED UNDER MORILL AND HOMESTEAD

model 1: teachers ~ jim crow, urbanization, disposession (child control?)

>>>>

---
title: "Settler colonial projects explain variation in US state formation in the early 20th c."
author: "Frank Edwards"
date: "last updated `r Sys.Date()`"
output: bookdown::html_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)

options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
agents<-read_csv("./data/temp_agents.csv")

## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)

dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)
```

```{r run-models, cache = T, include = F, message = F}
#source("models.R")
```

@@

Settler colonialism deeply and unevenly impacts the development of social policy systems. Settler colonial regimes engage in diverse forms of dispossession, extraction, exploitation, and social reproduction. We show that spatial and temporal variation in the salience of settler projects is closely related to the kinds of infrastructure that settler colonial states develop for social control and social reproduction. We show that the formation of the social state in the early 20th century in the US varied dramatically across places as a function of geographic variation in the salience of a set of interests of the settler colonial project: land dispossession, labor exploitation, and reproduction of the settler polity.

We examine three deeply intertwined, but spatially and temporally variable, contexts for local policymaking in the Progressive Era (1900 - 1930): the Jim Crow agricultural economy; mass urbanization and industrialization; and the rapid distribution of Native lands to white settlers. Each of these contexts presented different sets of problems to local and national settler policy elites.

Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests. Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital. In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers. White settlers and settler corporate interests flooded recently disposessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.

We suggest that each of these contexts yielded different trajectories for the formation of US state infrastructure. In the analysis that follows, we leverage data from a range of archival sources to map contexts for early 20th century settler colonization across the US, then build statistical models to evaluate whether these contexts were related to the kinds of capacities that local governments developed for contemporaneous social control and social reproduction.

Consistent with the historigraphy on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure. State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects. In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding shcools. In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies. Social control and reproduction instead relied on a coordinated formal and informal system of racial terror that developed alongside a relatively weak set of state social policy institutions.

The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US. States developed more or less robust institutions in various domains to address the problems of governance posed by the local projects of settler colonization that both elites and ordinary settlers sought to advance.

We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity. We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation. Or, put more directly, that THE LOCAL SALIENCE OF PARTICULAR CONSITUENTS / STRUCTURES IS RELATED TO STATE FORMATION.

These are three of the projects that were critical to settler colonialism in the early 20th c US. Not the only ones of course! Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILLIPINES COLONIZATION, AND OTHERS). *Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*

# Three settler colonial projects in the early 20th c. US

The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps). The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast. While Native land disposession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries. In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.

The Jim Crow system of apartheid was developed and becoming deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people. As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.

And the Industrial Revolution, coupled with a wave of crises across Eastern and Southern Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becomingly increasingly urban and foreign born.

Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'. During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels. Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.

```{r three-maps, fig.cap = "Geographic and demographic contexts in 1910 for settler colonial governance"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1910) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")

p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = blk / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "2. Black population")

p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "3. Urban immigrant population")

grid.arrange(p1, p2, p3, nrow = 1)
```

## Settling the frontier

OVERVIEW: BUILDING THE SETTLER NATION STATE, ERASING NATIVE NATIONS

Dispossession

Redistribution

THE MORILL, HOMESTEAD AND RR ACTS

Nation building

Forced assimilation

```{r fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

@@

ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year") 
```

@@

```{r fig.cap = "Disposession of Native lands in the 19th c."}
p1<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = cumulative_ceded / 1e6)) + 
  labs(x = "Year",
       y = "Acres (millions)") + 
  lims(y = c(0, 150)) + 
  facet_wrap(region~division) + 
  labs(subtitle = "1. Total area")

@@

```{r fig.cap = "Lands patented under two legislative mechanisms for the distribution of disposessed Native Lands to settlers, cuulative total 1930"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1930) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Morill Act") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Homestead Act") + 
  theme(legend.position = "bottom")

@@

## Entrenching the exploitation of Black labor and marginalization of Black social and political power

Between 1900 and 1930, settler elites in the the US South were busy entrenching Jim Crow and profiting from the deeply exploitative sharecropping system of agricultural production.

Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period. In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agrictultural production. In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.

### The sharecropping economy

DEFINE SYSTEM

SUMMARIZE LITERATURE

### The white southern polity

SUMMARIZE LIT ON SOUTHERN WHITENESS AND COHERENCE OF WHITE SETTLER POLITY IN US SOUTH

### The early and mid-career of Jim Crow

Racial order, law, and terrorism

ESP THINK ON BUREAUCRACY, INFORMALITY OF SOCIAL CONTROL, COLLABORATION BETWEEN FORMAL AND INFORMAL, VIGILANTISM AND LAW

EDUCATION AND JIM CROW

SOCIAL WORK AND JIM CROW

```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = blk_sharecrop / 1e5)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(hundred thousands)",
       subtitle = "1. Black population living on tenant farms") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / (white_immig + white_non_immig))) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Proportion",
        subtitle = "2. Proportion of white poulation foreign born") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)

@@

## Incorporating urban white immigrants into the settler polity and industrial working class

The industrialization of the US Northeast and Midwest during the late 19th and early 20th centuries spurred two related demographic processes: rapid urbanization and mass European immigration. Figure \@ref(fig:immig-maps) displays the distribution of both the urban population and the white immigrant population in the US in 1910. States like New York, Massachusetts, Pennsylvania, Ohio, and Illinois saw dramatic growth in their populations, with much of that growth composed of European immigrants and concentrated in cities like Chicago, Boston, Cleveland, Philadelphia, and New York.

```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
       subtitle = "1. White, foreign-born") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = urban_pop / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
        subtitle = "2. Urban") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)
```

SUMMARIZE DEMOGRAPHY LIT

### The rise of progressive domestic governance in Northern US cities

social work

policing

## Patterns in the development of state infrastructure 1900 - 1930

state capacity is hard to measure, but lets try! Here we are going to use employment measures for three roles that came to play critical roles in the developing practices of social welfare and social control: teachers, police, and social workers.

We also think about physical infrastructure for similar sets of institutions. We look at various forms of institutional group quarters: correctional institutions, mental institutions, and institutions for the poor and disabled.

Indian Boarding schools are another key piece of physical infrastructure. We look at those too, but I am still waiting on validating some of that data.

### Employment in education, policing, and social welfare: 1900 - 1930

There was dramatic growth in public sector employment in general (CHECK THIS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era. *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (CITES).*

In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930. Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase. Employment of police doubled between during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930. Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900). However, by 1930, the Census recorded about 17,000 social workers nationally, and increase of over 40 fold in terms of employment rates per capita.

```{r emp-ts, fig.cap = "Employment in three core social policy professions over time, US rates per 1,000 population, 1900 - 1930"}
emp_pdat<-pop %>% 
  group_by(year) %>% 
  summarize(pop = sum(tot_pop),
            Teachers = sum(teachers),
            Police = sum(police),
            social_workers = sum(socwork)) %>% 
  pivot_longer(cols = Teachers:social_workers) %>% 
  mutate(name = case_when(
    name == "Teachers"~ "1. Teachers",
    name == "Police" ~ "2. Police",
    name == "social_workers" ~ "3. Social Workers")) %>% 
  mutate(rate = value / pop * 1e3)

emp_pdat  %>% 
  ggplot(aes(x = year, y = rate)) + 
  geom_line() + 
  facet_wrap(~name, scales = "free_y") + 
  labs(x = "Year", y = "Employment rate")

However, this expansion of the reach of the state did not occur uniformly across the country. Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States. Employment rates for teachers were lowest-in-the nation in the US South. Employment rates for teachers were highest in a cluster of Great Plains states, like the Dakotas, Wyoming, Nebraska, Iowa, and Kansas. Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California. Police employment rates were relatively lower in places like Arkansas, Missisippi, and North Carolina. The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Missisippi and South Dakota recorded very few social workers in the 1930 Census.

```{r emp-maps, fig.cap = "Employment in three core social policy professions across US states in 1930, rates per 1,000 population"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1930)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = teachers / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
       subtitle = "1. Teachers") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = police / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "2. Police") + 
  theme(legend.position = "bottom")

p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = socwork / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "3. Social Workers") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, nrow = 1)
```

### Institutional capacity: prisons, jails, asylums, orphanages and poorhouses

State capacity could also be measured through the ability of states to engage in therapeutic, correctional, or paternalistic confinement. In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states. Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Using data from the Department of the Interior's ``Federal Indian Boarding School Initiative'' (2022), we display a count of Federal Indian boarding schools operating in 1930 in panel 4. I MAY ADD ENROLLMENT DATA FROM ARCIA REPORTS ONCE WE VALIDATE IT.

Data on institutional group quarters populations are less reliable than employment data for constructing a full 1900 - 1930 Census time series.

shows Prisons in NE, S, CA. Mental and poverty instutitons mostly in Urban NE, MW, CA

```{r gq-map, fig.cap = "Institutional group quarters, US 1930"}
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_correctional / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
       subtitle = "1. Correctional institutions") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_mental / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
        subtitle = "2. Mental institutions") + 
  theme(legend.position = "bottom")

p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_poor_disabled / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
        subtitle = "3. Institutions for the\npoor and disabled") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, nrow = 2)
```

### Institutional capacity: Indian boarding schools

FROM DOI REPORTS
